Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-12 paper.Rmd
Attempt to replace section 9 with a summary

All mathematical formulations, parameter specifications, and supporting figures are detailed in the Appendix.

@@

To achieve a near-optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner. In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose. A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.

The number of base filters determines the number of filters for the first 2D convolutional layer. In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one. This hyperparameter aids in controlling the complexity of the computer vision model. A higher number of base filters results in more trainable parameters. Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block. Increasing the number of units enhances model complexity, resulting in more trainable parameters.

The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training. A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout]. Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge. Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep]. It helps stabilize and accelerate the training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy. Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.

The learning rate is a crucial hyperparameter, as it dictates the step size of the optimization algorithm. A high learning rate can help the model avoid local minima but risks overshooting and missing the global minimum. Conversely, a low learning rate smoothens the training process but makes the convergence time longer and increases the likelihood of getting trapped in local minima.

Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras]. During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data. The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error. The tuner then restored the best epoch of the best model from the trials. Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs. The maximum allowed epochs was set at 2,000, although no models reached this threshold.

```{r}
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           Domain = c("\\{4, 8, 16, 32, 64\\}", "[0.1, 0.6]", "\\{false, true\\}", "\\{max, average\\}", "\\{false, true\\}", "\\{128, 256, 512, 1024, 2048\\}", "\\{false, true\\}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
  kableExtra::kable(format = "latex",
                    booktabs = TRUE, 
                    label = "hyperparameter",
                    caption = "Name of hyperparameters and their correspoding domain for the computer vision model.",
                    escape = FALSE)
```

Based on the tuning process described above, the optimized hyperparameter values are presented in Table \ref{tab:best-hyperparameter}. It was observable that a minimum of $32$ base filters was necessary, with the preferable choice being $64$ base filters for both the $64 \times 64$ and $128 \times 128$ models, mirroring the original VGG16 architecture. The optimized dropout rate for convolutional blocks hovered around $0.4$, and incorporating batch normalization for convolutional blocks proved beneficial for performance.

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex. The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.

```{r}
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           `32` = c("32", "0.4", "true", "max", "false", "256", "false", "0.2", "0.0003"),
           `64` = c("64", "0.3", "true", "average", "false", "256", "true", "0.4", "0.0006"),
           `128` = c("64", "0.4", "true", "average", "false", "256", "true", "0.1", "0.0052")) %>%
  kableExtra::kable(format = "latex",
                    booktabs = TRUE, 
                    label = "best-hyperparameter",
                    caption = "Hyperparameters values for the optimized computer vision models with different input sizes.",
                    col.names = c("Hyperparameter", "$32 \\times 32$", "$64 \\times 64$", "$128 \\times 128$"),
                    escape = FALSE) %>%
  kableExtra::kable_styling(latex_options = c("scale_down"))
```

>>>>

The mathematical formulations of the synthetic data model, parameter specifications, supporting tables and example figures of visual patterns of simulated residuals are provided in the Appendix.

@@

The computer vision models were trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras], with hyperparameters optimized via Bayesian tuning in `KerasTuner` [@omalley2019kerastuner] to minimize validation RMSE. The tuning process involved 100 trials, early stopping, and consideration of factors such as dropout, batch normalization, input resolution, and auxiliary inputs. Details of the hyperparameter domains, optimization strategy, and final selected values are provided in Appendix.

### 2025-06-12 paper.tex
Attempt to replace section 9 with a summary

To achieve a near-optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

The number of base filters determines the number of filters for the
first 2D convolutional layer. In the VGG16 architecture, the number of
filters for the 2D convolutional layer in a block is typically twice the
number in the previous block, except for the last block, which maintains
the same number of convolution filters as the previous one. This
hyperparameter aids in controlling the complexity of the computer vision
model. A higher number of base filters results in more trainable
parameters. Likewise, the number of units for the fully-connected layer
determines the complexity of the final prediction block. Increasing the
number of units enhances model complexity, resulting in more trainable
parameters.

The dropout rate and batch normalization are flexible hyperparameters
that work in conjunction with other parameters to facilitate smooth
training. A higher dropout rate is necessary when the model tends to
overfit the training dataset by learning too much noise
\citep{srivastava2014dropout}. Conversely, a lower dropout rate is
preferred when the model is complex and challenging to converge. Batch
normalization, on the other hand, addresses the internal covariate shift
problem arising from the randomness in weight initialization and input
data \citep{goodfellow2016deep}. It helps stabilize and accelerate the
training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and
the number of observations can potentially enhance prediction accuracy.
Therefore, we allow the tuner to determine whether these inputs were
necessary for optimal model performance.

The learning rate is a crucial hyperparameter, as it dictates the step
size of the optimization algorithm. A high learning rate can help the
model avoid local minima but risks overshooting and missing the global
minimum. Conversely, a low learning rate smoothens the training process
but makes the convergence time longer and increases the likelihood of
getting trapped in local minima.

Our model was trained on the MASSIVE M3 high-performance computing
platform \citep{goscinski2014multi}, using TensorFlow
\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
training, 80\% of the training data was utilized for actual training,
while the remaining 20\% was used as validation data. The Bayesian
optimization tuner conducted 100 trials to identify the best
hyperparameter values based on validation root mean square error. The
tuner then restored the best epoch of the best model from the trials.
Additionally, we applied early stopping, terminating the training
process if the validation root mean square error fails to improve for 50
epochs. The maximum allowed epochs was set at 2,000, although no models
reached this threshold.

\caption{\label{tab:hyperparameter}Name of hyperparameters and their correspoding domain for the computer vision model.}
\centering
\begin{tabular}[t]{ll}
\toprule
Hyperparameter & Domain\\
\midrule
Number of base filters & \{4, 8, 16, 32, 64\}\\
Dropout rate for convolutional blocks & {}[0.1, 0.6]\\
Batch normalization for convolutional blocks & \{false, true\}\\
Type of global pooling & \{max, average\}\\
Ignore additional inputs & \{false, true\}\\
\addlinespace
Number of units for the fully-connected layer & \{128, 256, 512, 1024, 2048\}\\
Batch normalization for the fully-connected layer & \{false, true\}\\
Dropout rate for the fully-connected layer & {}[0.1, 0.6]\\
Learning rate & {}[$10^{-8}$, $10^{-1}$]\\
\bottomrule
\end{tabular}
\end{table}

Based on the tuning process described above, the optimized
hyperparameter values are presented in Table
\ref{tab:best-hyperparameter}. It was observable that a minimum of
\(32\) base filters was necessary, with the preferable choice being
\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
models, mirroring the original VGG16 architecture. The optimized dropout
rate for convolutional blocks hovered around \(0.4\), and incorporating
batch normalization for convolutional blocks proved beneficial for
performance.

All optimized models chose to retain the additional inputs, contributing
to the reduction of validation error. The number of units required for
the fully-connected layer was \(256\), a relatively modest number
compared to the VGG16 classifier, suggesting that the problem at hand
was less complex. The optimized learning rates were higher for models
with higher resolution input, likely because models with more parameters
are more prone to getting stuck in local minima, requiring a higher
learning rate.

\begin{table}
\centering
\caption{\label{tab:best-hyperparameter}Hyperparameters values for the optimized computer vision models with different input sizes.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}[t]{llll}
\toprule
Hyperparameter & $32 \times 32$ & $64 \times 64$ & $128 \times 128$\\
\midrule
Number of base filters & 32 & 64 & 64\\
Dropout rate for convolutional blocks & 0.4 & 0.3 & 0.4\\
Batch normalization for convolutional blocks & true & true & true\\
Type of global pooling & max & average & average\\
Ignore additional inputs & false & false & false\\
\addlinespace
Number of units for the fully-connected layer & 256 & 256 & 256\\
Batch normalization for the fully-connected layer & false & true & true\\
Dropout rate for the fully-connected layer & 0.2 & 0.4 & 0.1\\
Learning rate & 0.0003 & 0.0006 & 0.0052\\
\bottomrule
\end{tabular}}
\end{table}

>>>>

The computer vision models were trained on the MASSIVE M3
high-performance computing platform \citep{goscinski2014multi}, using
TensorFlow \citep{abadi2016tensorflow} and Keras
\citep{chollet2015keras}, with hyperparameters optimized via Bayesian
tuning in \texttt{KerasTuner} \citep{omalley2019kerastuner} to minimize
validation RMSE. The tuning process involved 100 trials, early stopping,
and consideration of factors such as dropout, batch normalization, input
resolution, and auxiliary inputs. Details of the hyperparameter domains,
optimization strategy, and final selected values are provided in
Appendix.
